---
title: "STATS 206 Pset 2"
author: "Léon Marbach"
date: "2023-10-17"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/lmarbach/Library/CloudStorage/GoogleDrive-lmarbach@stanford.edu/My Drive/Coursework/Year 2/Fall 2023/STATS 206/pset2")
```

# Problem 1

## (a)

```{r}
# create matrices
sigma_A <- matrix(c(2.929, 5.898, 5.386, 
                    5.898, 19.032, 16.467, 
                    5.386, 16.467, 16.038),
                  nrow = 3, ncol = 3) 
sigma_B <- matrix(c(30.077, -4.956, -7.673, 
                    -4.956, 20.443, -15.085, 
                    -7.673, -15.085, 19.480),
                  nrow = 3, ncol = 3) 
sigma_C <- matrix(c(2.384, 0.173, -0.164, 
                    0.173, 2.091, -0.006, 
                    -0.164, -0.006, 2.425),
                  nrow = 3, ncol = 3) 

# find eigenvalues and eigenvectors
eigen(sigma_A)
eigen(sigma_B)
eigen(sigma_C)
```

We know that the eigenvalues we found for each multivariate normal distribution determine the lengths of semiaxes of the 3-dimensional ellipsoid (stretching along each direction). Hence, distribution A has the shape of a **needle** (one eigenvalue much larger than the other two that are approximately the same), distribution B has the shape of a **plate** (two eigenvalues much larger than the third one - note that the plate is not perfectly round because the two largest eigenvalues are not exactly the same), and distribution C has approximately the shape of a **ball** (three eigenvalues are around 2.3 - note that the ball is not perfectly round because eigenvalues are not exactly the same).

## (b)

The eigenvalues are always returned in decreasing order, and each column of vectors corresponds to the elements in values. Thus, the eigenvector that goes with the largest eigenvalue is the **first column vector**:

* for distribution A, the eigenvector that goes with the largest eigenvalue 35.9990069 is (-0.2347735, -0.7177842, -0.6554901)
* for distribution B, the eigenvector that goes with the largest eigenvalue 35.999962 is (-0.4652949, -0.5353195, 0.7049352)
* for distribution C, the eigenvector that goes with the largest eigenvalue 2.600198 is (0.7032421, 0.2467829, -0.6667449)

## (c)

See Appendix for details on calculations.

```{r}
# distrib A: variance of X2 conditionally on X1 and X3
sigma_A[2,2] - 
  matrix(c(sigma_A[2,1], sigma_A[2,3]), nrow = 1, ncol = 2) %*% 
  solve(matrix(c(sigma_A[1,1], sigma_A[3,1], sigma_A[1,3], sigma_A[3,3]), nrow = 2, ncol = 2)) %*%
  matrix(c(sigma_A[2,1], sigma_A[2,3]), nrow = 2, ncol = 1)

# distrib B: correlation of X2 and X3 conditionally on X1
var_cov <- matrix(c(sigma_B[2,2], sigma_B[2,3], sigma_B[3,2], sigma_B[3,3]), nrow = 2, ncol = 2) -
    (1 / sigma_B[1,1]) *
    matrix(c(sigma_B[2,1], sigma_B[3,1]), nrow = 2, ncol = 1) %*%
    matrix(c(sigma_B[2,1], sigma_B[3,1]), nrow = 1, ncol = 2)
var_cov[1,2] / sqrt(var_cov[1,1] * var_cov[2,2])

# distrib C: expectation of X3 conditionally on X1 = 1 and X2 = -1
matrix(c(sigma_C[3,1], sigma_C[3,2]), nrow = 1, ncol = 2) %*%
  solve(matrix(c(sigma_C[1,1], sigma_C[1,2], sigma_C[1,2], sigma_C[2,2]), nrow = 2, ncol = 2)) %*%
  matrix(c(1, -1), nrow = 2, ncol = 1)
```

# Problem 2

## (a)

```{r}
# import datasets
male <- read.csv("ANSUR_II_MALE_Public.csv")
female <- read.csv("ANSUR_II_FEMALE_Public.csv")

# select out dimensions containing length
male_length <- male[, grep("length", names(male))]
female_length <- female[, grep("length", names(female))]

# compute T2 statistic
library(DescTools)
HotellingsT2Test(male_length, female_length)
```
The $T^2$ statistic based on the F-distribution is 873.23.

## (b)

```{r}
# divide all columns by height
male_length_normalized <- male_length/male$Heightin
female_length_normalized <- female_length/female$Heightin

# compute T2 statistic
HotellingsT2Test(male_length_normalized, female_length_normalized)
```

The $T^2$ statistic based on the F-distribution is 552.65.

## (c)

```{r}
# select length variables in male dataset and separate by writing preference
male_length_left <- subset(male[, grep("length", names(male))], 
                           male$WritingPreference == "Left hand")
male_length_right <- subset(male[, grep("length", names(male))], 
                            male$WritingPreference == "Right hand")

# compute T2 statistic
HotellingsT2Test(male_length_left, male_length_right)
```

For men, the $T^2$ statistic based on the F-distribution is 2.5709 and the p-value is 0.0001071.

```{r}
# select length variables in female dataset and separate by writing preference
female_length_left <- subset(female[, grep("length", names(female))], 
                           female$WritingPreference == "Left hand")
female_length_right <- subset(female[, grep("length", names(female))], 
                              female$WritingPreference == "Right hand")

# compute T2 statistic
HotellingsT2Test(female_length_left, female_length_right)
```

For women, the $T^2$ statistic based on the F-distribution is 0.92602 and the p-value is 0.5567.

## (d)

```{r}
# 1. male vs female

# select out dimensions containing breadth
male_breadth <- male[, grep("breadth", names(male))]
female_breadth <- female[, grep("breadth", names(female))]

# compute T2 statistic
HotellingsT2Test(male_breadth, female_breadth)
```

The $T^2$ statistic based on the F-distribution is 1991.2.

```{r}
# 2. normalized male vs female

# divide all columns by weight
# remove id 13956 who has weight 0
male_subset <- male[male$subjectid != 13956, ]
male_breadth_subset <- male_subset[, grep("breadth", names(male))]
male_breadth_normalized <- male_breadth_subset/male_subset$Weightlbs
female_breadth_normalized <- female_breadth/female$Weightlbs

# compute T2 statistic
HotellingsT2Test(male_breadth_normalized, female_breadth_normalized)
```

The $T^2$ statistic based on the F-distribution is 2233.8.

```{r}
# 3a. left vs right-handed male

# select breadth variables in male dataset and separate by writing preference
male_breadth_left <- subset(male[, grep("breadth", names(male))], 
                           male$WritingPreference == "Left hand")
male_breadth_right <- subset(male[, grep("breadth", names(male))], 
                             male$WritingPreference == "Right hand")

# compute T2 statistic
HotellingsT2Test(male_breadth_left, male_breadth_right)
```

For men, the $T^2$ statistic based on the F-distribution is 3.2229 and the p-value is 1.405e-05.

```{r}
# 3b. left vs right-handed female

# select breadth variables in female dataset and separate by writing preference
female_breadth_left <- subset(female[, grep("breadth", names(female))], 
                           female$WritingPreference == "Left hand")
female_breadth_right <- subset(female[, grep("breadth", names(female))], 
                            female$WritingPreference == "Right hand")

# compute T2 statistic
HotellingsT2Test(female_breadth_left, female_breadth_right)
```

For women, the $T^2$ statistic based on the F-distribution is 1.0719 and the p-value is 0.3768.

# Problem 3

## (a)

```{r}
# import iris dataset
data(iris)
pairs(iris[,1:4],col=as.numeric(iris[,5]))

# print variance-covariance matrix with 3 significant figures
# setosa
cov_setosa <- signif(cov(iris[iris$Species == "setosa", 1:4]), digits = 3)
cov_setosa
# versicolor
cov_versicolor <- signif(cov(iris[iris$Species == "versicolor", 1:4]), digits = 3)
cov_versicolor
# virginica
cov_virginica <- signif(cov(iris[iris$Species == "virginica", 1:4]), digits = 3)
cov_virginica
```

## (b)

```{r}
# print means for each species
# setosa
sapply(iris[iris$Species == "setosa", 1:4], mean)
# versicolor
sapply(iris[iris$Species == "versicolor", 1:4], mean)
# virginica
sapply(iris[iris$Species == "virginica", 1:4], mean)
```

## (c)

```{r}
# pooled variance
S <- ((50 - 1) * cov_virginica + (50 - 1) * cov_versicolor) / (50 + 50 - 2)
S

# compute T2 statistic
HotellingsT2Test(iris[iris$Species == "versicolor", 1:4], iris[iris$Species == "virginica", 1:4])
```

The p-value is < 2.2e-16.

## (d)

```{r}
# between groups SSCP matrix B
means <- by(iris[, 1:4], iris$Species, colMeans)
overall_mean <- colMeans(iris[, 1:4])
B <- matrix(0, ncol = 4, nrow = 4)
for (i in 1:3) {
  B <- B + nrow(iris[iris$Species == names(means)[i],]) * 
       (means[[i]] - overall_mean) %*% t(means[[i]] - overall_mean)
}
B

# within groups (pooled) SSCP W
W <- matrix(0, ncol = 4, nrow = 4)
for (i in 1:3) {
  subdata <- iris[iris$Species == names(means)[i], 1:4]
  W <- W + (nrow(subdata) - 1) * var(subdata)
}
W

# Wilks' lambda
lambda <- det(W)/det(B+W)
lambda

# test statistic
test_statistic <- - (150 - 1 - (4 + 3) / 2) * log(lambda)
test_statistic

# p value
p_value <- pchisq(q = test_statistic, df = 4 * (3 - 1), lower.tail = F)
p_value
```

## (e)

```{r}
# between groups SSCP matrix B
means <- by(iris[, 3:4], iris$Species, colMeans)
overall_mean <- colMeans(iris[, 3:4])
B <- matrix(0, ncol = 2, nrow = 2)
for (i in 1:3) {
  B <- B + nrow(iris[iris$Species == names(means)[i],]) * 
       (means[[i]] - overall_mean) %*% t(means[[i]] - overall_mean)
}
B

# within groups (pooled) SSCP W
W <- matrix(0, ncol = 2, nrow = 2)
for (i in 1:3) {
  subdata <- iris[iris$Species == names(means)[i], 3:4]
  W <- W + (nrow(subdata) - 1) * var(subdata)
}
W

# Wilks' lambda
lambda <- det(W)/det(B+W)
lambda

# test statistic
test_statistic <- - (150 - 1 - (2 + 3) / 2) * log(lambda)
test_statistic

# p value
p_value <- pchisq(q = test_statistic, df = 2 * (3 - 1), lower.tail = F)
p_value
```
