---
title: "STATS 206 Pset 3"
author: "Léon Marbach"
date: "2023-10-24"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/lmarbach/Library/CloudStorage/GoogleDrive-lmarbach@stanford.edu/My Drive/Coursework/Year 2/Fall 2023/STATS 206/pset3")
```

# Problem 1

```{r}
# import datasets
male <- read.csv("ANSUR_II_MALE_Public.csv")
female <- read.csv("ANSUR_II_FEMALE_Public.csv")

# keep first 93 variables, except subjectid (1st variable)
male_subset <- male[, 2:94]
female_subset <- female[, 2:94]

# remove variable weightkg
male_subset <- subset(male_subset, select = -c(weightkg))
female_subset <- subset(female_subset, select = -c(weightkg))

# PCA on correlation matrix
pca_male <- prcomp(male_subset, scale.=T)
pca_female <- prcomp(female_subset, scale.=T)

# extract coefficients of the second PC loading
coeff_table_male <- data.frame(variable = colnames(male_subset), 
                               coefficient = pca_male$rotation[, 2])
coeff_table_female <- data.frame(variable = colnames(female_subset), 
                                 coefficient = pca_female$rotation[, 2])

# Sort the table by the absolute values of the coefficients
coeff_table_male <- coeff_table_male[order(abs(coeff_table_male$coefficient), 
                                           decreasing = TRUE), ]
coeff_table_female <- coeff_table_female[order(abs(coeff_table_female$coefficient), 
                                               decreasing = TRUE), ]

# Print the sorted coefficient tables
print(coeff_table_male, row.names = FALSE)
print(coeff_table_female, row.names = FALSE)

# male: smallest 5 coeff of 2nd PC eigenvector
# head(sort(pca_male$rotation[, 2]), 5)
# male: largest 5 coeff of 2nd PC eigenvector
# tail(sort(pca_male$rotation[, 2]), 5)
# female: smallest 5 coeff of 2nd PC eigenvector
# head(sort(pca_female$rotation[, 2]), 5)
# female: largest 5 coeff of 2nd PC eigenvector
# tail(sort(pca_female$rotation[, 2]), 5)
```

I am printing the full table of coefficients instead of just the top 5 or bottom 5 to see both the top and bottom in absolute value (values close to zero and values far from zero) - the code to print the largest and smallest 5 (no absolute value) still appears as comments.

The second PC eigenvector represents the direction in the data that captures the most variance after accounting for the first PC. It is orthogonal to the first PC. Weights far from zero indicate that the variable strongly influences the component, while weights close to 0 indicate that the variable has a weak influence on the component. We can also look at weight signs to see whether a variable and a PC are positively or negatively correlated. 

Interpretation is not obvious! For both male and female datasets, the largest positive coefficients represent how fat or big the belly/waist/chest of a person is ("waist circumference", "abdominal extension depth sitting", "waist depth", ...) while the largest negative coefficients look more at the height of the person ("popliteal height", "crotch height",  "waist height omphalion", ...), suggesting that taller people are more slim and shorter people are more fat (which is expected). The second PC thus captures this morphological pattern / variability in girth versus height of the person.

# Problem 2

```{r}
# get scores on first three PC
male_scores <- pca_male$x[, 1:3]
female_scores <- pca_female$x[, 1:3]

# fit linear regression model
# male
lm_male <- lm(male$Age ~ male_scores[, 1] + male_scores[, 2] + male_scores[, 3])
summary(lm_male)
# female
lm_female <- lm(female$Age ~ female_scores[, 1] + female_scores[, 2] + female_scores[, 3])
summary(lm_female)
```

# Problem 3

```{r}
# anova male branch / PC1
aov_male_PC1 <- aov(male_scores[, 1] ~ male$Branch)
summary(aov_male_PC1)

# anova male branch / PC2
aov_male_PC2 <- aov(male_scores[, 2] ~ male$Branch)
summary(aov_male_PC2)

# anova male branch / PC3
aov_male_PC3 <- aov(male_scores[, 3] ~ male$Branch)
summary(aov_male_PC3)

# anova female branch / PC1
aov_female_PC1 <- aov(female_scores[, 1] ~ female$Branch)
summary(aov_female_PC1)

# anova female branch / PC2
aov_female_PC2 <- aov(female_scores[, 2] ~ female$Branch)
summary(aov_female_PC2)

# anova female branch / PC3
aov_female_PC3 <- aov(female_scores[, 3] ~ female$Branch)
summary(aov_female_PC3)
```

# Problem 4

## (a)

```{r}
# import dataset
sevens = as.matrix(read.table("sevens.txt"))

# see what images look like
image(matrix(sevens[1, ], 28, 28))
image(matrix(sevens[1, ], 28, 28)[, 28:1])

# compute principal components
prcomp_sevens <- prcomp(sevens)

# PC1
image(matrix(prcomp_sevens$rotation[, 1], 28, 28)[,28:1])

# PC2
image(matrix(prcomp_sevens$rotation[, 2], 28, 28)[,28:1])

# PC3
image(matrix(prcomp_sevens$rotation[, 3], 28, 28)[,28:1])

# PC4
image(matrix(prcomp_sevens$rotation[, 4], 28, 28)[,28:1])

# PC5
image(matrix(prcomp_sevens$rotation[, 5], 28, 28)[,28:1])

# PC6
image(matrix(prcomp_sevens$rotation[, 6], 28, 28)[,28:1])

# PC7
image(matrix(prcomp_sevens$rotation[, 7], 28, 28)[,28:1])

# PC8
image(matrix(prcomp_sevens$rotation[, 8], 28, 28)[,28:1])

# PC9
image(matrix(prcomp_sevens$rotation[, 9], 28, 28)[,28:1])
```

## (b)

```{r}
# calculate proportion of variance explained by each principal component
prop_var_explained <- prcomp_sevens$sdev^2 / sum(prcomp_sevens$sdev^2)

# create cumulative proportion of variance explained vector
cumulative_prop_var <- cumsum(prop_var_explained)

# create plot to show proportion of variance explained by number of principal components
plot(1:length(prop_var_explained), cumulative_prop_var, type = "b", 
     xlab = "Number of Principal Components", ylab = "Proportion of Variance Explained",
     main = "Proportion of Variance Explained by Principal Components")

# find how many principal components needed to get 90% of the variance
which(cumulative_prop_var >= 0.9)[1]
```

We need 66 PC to get 90% of the variance.