---
title: "STATS 206 Pset 4"
author: "Léon Marbach"
date: "2023-11-16"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/lmarbach/Library/CloudStorage/GoogleDrive-lmarbach@stanford.edu/My Drive/Coursework/Year 2/Fall 2023/STATS 206/pset4")
```

```{r}
set.seed(20231107)
```

# Problem 1

## (a)

```{r}
makevals = function( N = 2500, n = 10, d = 2, seed = 20231107 ){
set.seed(seed)
usa = matrix( rnorm(N*d),nrow=N )
hi = matrix( rnorm(n*d),nrow=n )*.2
hi[,1] = hi[,1]-6
hi[,2] = hi[,2]-3
all = rbind(usa,hi)
all = round(all,2)
gp = c(rep(1,N),rep(2,n))
colnames(all)=c("long","lat")
plot( all[,1],all[,2],xlab="Longitude",ylab="Lattitude",
main="Data for k means", col=gp, pch=16, asp=1)
all
}

data <- makevals()

kmeans_a <- kmeans(data, centers = 2)

points(
  data[, 1], data[, 2], 
  col = ifelse(kmeans_a$cluster == 1, "green", "yellow"),
  pch = 16, 
  cex = 1
)

points(
  kmeans_a$centers[, 1], kmeans_a$centers[, 2],
  col = "black", 
  pch = 3, 
  cex = 3
)

legend(
  "topright", 
  legend = c("Cluster 1", "Cluster 2", "Cluster centers"), 
  col = c("green", "yellow", "black"),
  pch = c(16, 16, 3), 
  cex = 0.8
)
```

It does not find Hawaii.

## (b)

```{r}
makevals = function( N = 2500, n = 10, d = 2, seed = 20231107 ){
set.seed(seed)
usa = matrix( rnorm(N*d),nrow=N )
hi = matrix( rnorm(n*d),nrow=n )*.2
hi[,1] = hi[,1]-6
hi[,2] = hi[,2]-3
all = rbind(usa,hi)
all = round(all,2)
gp = c(rep(1,N),rep(2,n))
colnames(all)=c("long","lat")
plot( all[,1],all[,2],xlab="Longitude",ylab="Lattitude",
main="Data for k means", col=gp, pch=16, asp=1)
all
}

data <- makevals()

kmeans_b_1 <- kmeans(data, centers = 2)
kmeans_b_1$centers
kmeans_b_1$iter

kmeans_b_2 <- kmeans(data, centers = 2)
kmeans_b_2$centers
kmeans_b_2$iter

kmeans_b_3 <- kmeans(data, centers = 2)
kmeans_b_3$centers
kmeans_b_3$iter

kmeans_b_4 <- kmeans(data, centers = 2)
kmeans_b_4$centers
kmeans_b_4$iter

kmeans_b_5 <- kmeans(data, centers = 2)
kmeans_b_5$centers
kmeans_b_5$iter

kmeans_b_6 <- kmeans(data, centers = 2)
kmeans_b_6$centers
kmeans_b_6$iter

kmeans_b_7 <- kmeans(data, centers = 2)
kmeans_b_7$centers
kmeans_b_7$iter

kmeans_b_8 <- kmeans(data, centers = 2)
kmeans_b_8$centers
kmeans_b_8$iter

kmeans_b_9 <- kmeans(data, centers = 2)
kmeans_b_9$centers
kmeans_b_9$iter

kmeans_b_10 <- kmeans(data, centers = 2)
kmeans_b_10$centers
kmeans_b_10$iter
```

When repeating the process 10 times, we observe that the number of iterations kmeans use is always 1. As for the cluster centers, we observe that either the cluster centers will be $(0.7060717,  0.05617625)$ and $(-0.9302135, -0.09851409)$, or $(0.7049142,  0.05538404)$ and $(-0.9316852, -0.09786997)$, which is very close.

## (c)

```{r}
makevals = function( N = 2500, n = 10, d = 2, seed = 20231107 ){
set.seed(seed)
usa = matrix( rnorm(N*d),nrow=N )
hi = matrix( rnorm(n*d),nrow=n )*.2
hi[,1] = hi[,1]-6
hi[,2] = hi[,2]-3
all = rbind(usa,hi)
all = round(all,2)
gp = c(rep(1,N),rep(2,n))
colnames(all)=c("long","lat")
plot( all[,1],all[,2],xlab="Longitude",ylab="Lattitude",
main="Data for k means", col=gp, pch=16, asp=1)
all
}

data <- makevals()

kmeans_c <- kmeans(data, centers = 3)

cluster_colors <- c("yellow", "green", "cyan")

points(
  data[, 1], data[, 2], 
  col = cluster_colors[kmeans_c$cluster],
  pch = 16, 
  cex = 1
)

points(
  kmeans_c$centers[, 1], kmeans_c$centers[, 2],
  col = "black", 
  pch = 3, 
  cex = 3
)

legend(
  "topright", 
  legend = c("Cluster 1", "Cluster 2", "Cluster 3", "Cluster centers"), 
  col = c("yellow", "green", "cyan", "black"),
  pch = c(16, 16, 16, 3), 
  cex = 0.8
)

kmeans_c_1 <- kmeans(data, centers = 3)
kmeans_c_1$centers
kmeans_c_1$iter

kmeans_c_2 <- kmeans(data, centers = 3)
kmeans_c_2$centers
kmeans_c_2$iter

kmeans_c_3 <- kmeans(data, centers = 3)
kmeans_c_3$centers
kmeans_c_3$iter

kmeans_c_4 <- kmeans(data, centers = 3)
kmeans_c_4$centers
kmeans_c_4$iter

kmeans_c_5 <- kmeans(data, centers = 3)
kmeans_c_5$centers
kmeans_c_5$iter

kmeans_c_6 <- kmeans(data, centers = 3)
kmeans_c_6$centers
kmeans_c_6$iter

kmeans_c_7 <- kmeans(data, centers = 3)
kmeans_c_7$centers
kmeans_c_7$iter

kmeans_c_8 <- kmeans(data, centers = 3)
kmeans_c_8$centers
kmeans_c_8$iter

kmeans_c_9 <- kmeans(data, centers = 3)
kmeans_c_9$centers
kmeans_c_9$iter

kmeans_c_10 <- kmeans(data, centers = 3)
kmeans_c_10$centers
kmeans_c_10$iter
```

Again, it does not find Hawaii.

When repeating the process 10 times, we observe that the number of iterations kmeans use is between 2 and 4. As for the cluster centers, they vary greatly and are systematically different.

## (d)

```{r}
makevals = function( N = 2500, n = 10, d = 2, seed = 20231107 ){
set.seed(seed)
usa = matrix( rnorm(N*d),nrow=N )
hi = matrix( rnorm(n*d),nrow=n )*.2
hi[,1] = hi[,1]-6
hi[,2] = hi[,2]-3
all = rbind(usa,hi)
all = round(all,2)
gp = c(rep(1,N),rep(2,n))
colnames(all)=c("long","lat")
plot( all[,1],all[,2],xlab="Longitude",ylab="Lattitude",
main="Data for k means", col=gp, pch=16, asp=1)
all
}

data <- makevals()

kmeans_d <- kmeans(data, centers = 7)

cluster_colors <- c(1:7)

points(
  data[, 1], data[, 2], 
  col = cluster_colors[kmeans_d$cluster],
  pch = 16, 
  cex = 1
)

points(
  kmeans_d$centers[, 1], kmeans_d$centers[, 2],
  col = "white",
  pch = 3,
  cex = 2
)

legend(
  "topright", 
  legend = c("Cluster 1", "Cluster 2", "Cluster 3", "Cluster 4", "Cluster 5",
             "Cluster 6", "Cluster 7", "Cluster centers"), 
  col = c(1:7, "black"),
  pch = c(16, 16, 16, 16, 16, 16, 16, 3), 
  cex = 0.8
)
```

The first time we get exactly one cluster in Hawaii is at $k=7$.

## (e)

```{r}
library(deldir)
cluster_centers <- kmeans_d$centers
cluster_centers_df <- data.frame(xloc = cluster_centers[, 1], yloc = cluster_centers[, 2])
tesselation <- deldir(cluster_centers_df$xloc, cluster_centers_df$yloc)
tiles <- tile.list(tesselation)
plot(tiles, pch=19)
```

# Problem 2

# Problem 3

```{r}
# import dataset
pop <- read.table("post-operative.data", header = F, sep = ",")

# rename variables
names(pop) <- c("int_temp", "surf_temp", "oxygen", "bp", "stab_surf_temp", 
                "stab_core_temp", "stab_bp", "comfort", "decision")

# handle NAs in comfort level
pop$comfort[pop$comfort == "?"] <- NA

# convert to factors
pop$int_temp <- factor(pop$int_temp, ordered = TRUE, levels = c("low", "mid", "high"))
pop$surf_temp <- factor(pop$surf_temp, ordered = TRUE, levels = c("low", "mid", "high"))
pop$oxygen <- factor(pop$oxygen, ordered = TRUE, levels = c("good", "excellent"))
pop$bp <- factor(pop$bp, ordered = TRUE, levels = c("low", "mid", "high"))
pop$stab_surf_temp <- factor(pop$stab_surf_temp, ordered = TRUE, 
                             levels = c("unstable", "stable"))
pop$stab_core_temp <- factor(pop$stab_core_temp, ordered = TRUE, 
                             levels = c("unstable", "mod-stable", "stable"))
pop$comfort <- as.numeric(pop$comfort)
pop$decision <- factor(pop$decision, ordered = FALSE, levels = c("I", "S", "A"))
```

3 observations have a missing value for the $comfort$ variable. Since one observation is an intensive care decision and we only have 2 intensive care decisions in our dataset, we decide not to remove this observation and the 2 others but to replace the three missing values with the mean of $comfort$ variable rounded to the nearest integer.

```{r}
# replace NAs with comfort mean
mean_comfort <- round(mean(pop$comfort, na.rm = TRUE))
pop$comfort[is.na(pop$comfort)] <- mean_comfort

# Load necessary packages
library(randomForest)
library(rpart)
library(caret)

# Train Decision Tree classifier on the full dataset
full_model_tree <- rpart(decision ~ ., data = pop, method = "class")

# Predict on the full dataset using Decision Tree
predictions_full_tree <- predict(full_model_tree, newdata = pop, type = "class")
conf_matrix_full_tree <- confusionMatrix(predictions_full_tree, pop$decision)

# Print Confusion Matrix for Decision Tree on Full Dataset
print(conf_matrix_full_tree)

# Train Random Forest classifier on the full dataset
full_model_rf <- randomForest(decision ~ ., data = pop, ntree = 500)

# Predict on the full dataset using Random Forest
predictions_full_rf <- predict(full_model_rf, newdata = pop, type = "response")
conf_matrix_full_rf <- confusionMatrix(predictions_full_rf, pop$decision)

# Print Confusion Matrix for Random Forest on Full Dataset
print(conf_matrix_full_rf)

# Perform 5-fold CV for Decision Tree
ctrl <- trainControl(method = "cv", number = 5)
cv_model_tree <- train(decision ~ ., data = pop, method = "rpart", trControl = ctrl)

# Print the results of cross-validation for Decision Tree
print(cv_model_tree$results)

# Perform 5-fold CV for Random Forest
ctrl <- trainControl(method = "cv", number = 5)
cv_model_rf <- train(decision ~ ., data = pop, method = "rf", trControl = ctrl)

# Print the results of cross-validation for Random Forest
print(cv_model_rf$results)
```

# Problem 4

## (a)

```{r}
# read data file
food <- read.table("food.txt", header = TRUE, row = 1)

# abbreviate variables
names(food) <- c("Fat", "Cal", "Carbs", "Protein", "Chol", "Weight", "SatFat")

# keep variable 6 for later
food_weight <- food$Weight
food <- food[, -which(names(food) == "Weight")]

# convert to matrix
food <- as.matrix(food)

# Scale the data
scaled_food <- scale(food)

# Hierarchical clustering with single linkage
hclust_single <- hclust(dist(scaled_food), method = "single")

# Hierarchical clustering with average linkage
hclust_average <- hclust(dist(scaled_food), method = "average")

# Hierarchical clustering with complete linkage
hclust_complete <- hclust(dist(scaled_food), method = "complete")

# plot dendrograms
par(mfrow = c(1, 3))
plot(hclust_single, main = "Single Linkage", xlab = "", ylab = "", labels = F)
plot(hclust_average, main = "Average Linkage", xlab = "", ylab = "", labels = F)
plot(hclust_complete, main = "Complete Linkage", xlab = "", ylab = "", labels = F)
```

## (b)

```{r}
# Cut the dendrograms to create clusters for each linkage method
cut_single <- cutree(hclust_single, k = 10)
cut_average <- cutree(hclust_average, k = 10)
cut_complete <- cutree(hclust_complete, k = 10)

# Create tables with the counts for each cluster
table_single <- table(cut_single)
table_average <- table(cut_average)
table_complete <- table(cut_complete)

# Combine the tables into a single data frame
cluster_table <- data.frame(
  Cluster = 1:10,
  Single = table_single,
  Average = table_average,
  Complete = table_complete
)

# Reshape the data frame to have methods in rows and clusters in columns
cluster_table_reshaped <- as.data.frame(t(cluster_table[, -1]))

# Print only the frequency rows
print(cluster_table_reshaped[grepl(".Freq", rownames(cluster_table_reshaped)), ])
```

## (c)

The clustering most affected by this is the single linkage clustering. The cluster least affected by this is the complete linkage clustering.

```{r}
# Set the desired number of foods for filtering
min_foods = 2
max_foods = 50

# Choose the complete linkage clustering result
clustering_result <- cutree(hclust_complete, k = 10)

# Find clusters with more than one food and fewer than 50 foods
selected_clusters <- which(table(clustering_result) > min_foods & table(clustering_result) < max_foods)

# List the foods in the selected clusters line by line
for (cluster_id in selected_clusters) {
  foods_in_cluster <- names(which(clustering_result == cluster_id))
  
  # Print cluster header
  cat("Cluster", cluster_id, ":\n")
  
  # Print each food on a new line
  for (food in foods_in_cluster) {
    cat("  - ", food, "\n")
  }
  cat("\n")
}
```

I pick clusters 6 and . Cluster 6 clearly brings together all oils. 

## (d)

```{r}
# read data file
food <- read.table("food.txt", header = TRUE, row = 1)

# abbreviate variables
names(food) <- c("Fat", "Cal", "Carbs", "Protein", "Chol", "Weight", "SatFat")

# keep variable 6 for later
food_weight <- food$Weight
food <- food[, -which(names(food) == "Weight")]

# convert to matrix
food <- as.matrix(food)

# divide by weight
food <- food / food_weight

# Scale the data
scaled_food <- scale(food)

# Hierarchical clustering with single linkage
hclust_single <- hclust(dist(scaled_food), method = "single")

# Hierarchical clustering with average linkage
hclust_average <- hclust(dist(scaled_food), method = "average")

# Hierarchical clustering with complete linkage
hclust_complete <- hclust(dist(scaled_food), method = "complete")

# plot dendrograms
par(mfrow = c(1, 3))
plot(hclust_single, main = "Single Linkage", xlab = "", ylab = "", labels = F)
plot(hclust_average, main = "Average Linkage", xlab = "", ylab = "", labels = F)
plot(hclust_complete, main = "Complete Linkage", xlab = "", ylab = "", labels = F)

# Cut the dendrograms to create clusters for each linkage method
cut_single <- cutree(hclust_single, k = 10)
cut_average <- cutree(hclust_average, k = 10)
cut_complete <- cutree(hclust_complete, k = 10)

# Create tables with the counts for each cluster
table_single <- table(cut_single)
table_average <- table(cut_average)
table_complete <- table(cut_complete)

# Combine the tables into a single data frame
cluster_table <- data.frame(
  Cluster = 1:10,
  Single = table_single,
  Average = table_average,
  Complete = table_complete
)

# Reshape the data frame to have methods in rows and clusters in columns
cluster_table_reshaped <- as.data.frame(t(cluster_table[, -1]))

# Print only the frequency rows
print(cluster_table_reshaped[grepl(".Freq", rownames(cluster_table_reshaped)), ])
```

Again, the most affected clustering is the single linkage clustering and the least affected is the complete linkage clustering.

## (e)

