---
title: "STATS 206 Pset 1"
author: "Léon Marbach"
date: "2023-10-03"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/lmarbach/Library/CloudStorage/GoogleDrive-lmarbach@stanford.edu/My Drive/Coursework/Year 2/Fall 2023/STATS 206/pset1")
```

# Problem 1: Food data

## (a)

```{r}
# read data file
food <- read.table("food.txt", header = TRUE)
# abbreviate variables
names(food) <- c("Fat", "Cal", "Carbs", "Protein", "Chol", "Weight", "SatFat")

correlation_matrix <- cor(food) # calculate correlation matrix
rounded_correlation_matrix <- round(correlation_matrix, 2) # round to two places
print(rounded_correlation_matrix) # print rounded correlation matrix
```

## (b)

```{r}
pairs(food) # print scatterplot matrix
print(food[which.max(food[["Chol"]]), ]) # find food with max cholesterol
print(food[which.max(food[["Fat"]]), ]) # find food with max fat
```

We can see that one food has a very high level of cholesterol compared to others: it is cheesecake (*CHEESECAKE 1 CAKE*). Another example of an extreme food is carrot cake (*CARROT CAKE,CREMCHESE FRST,REC1 CAKE*), which has a very high level of fat compared to other foods.

## (c)

```{r}
# divide each variable by Weight
food$Fat <- food$Fat/food$Weight
food$Cal <- food$Cal/food$Weight
food$Carbs <- food$Carbs/food$Weight
food$Protein <- food$Protein/food$Weight
food$Chol <- food$Chol/food$Weight
food$SatFat <- food$SatFat/food$Weight

pairs(food[, names(food) != "Weight"]) # print scatterplot matrix

print(food[which.max(food[["Chol"]]), ]) # find food with max cholesterol
print(food[which.max(food[["Protein"]]), ]) # find food with max protein
```
Normalizing by Weight makes new foods stand out. For example, we now see a food with extremely high level of cholesterol but it is not cheesecake anymore but eggs (*EGGS, RAW, YOLK 1 YOLK*). Another example that now stands out is a food with a high level of protein compared to others: gelatin (*GELATIN, DRY 1 ENVELP*).

# Problem 2: Mandibles

## (a)

```{r}
mandibles <- read.table("mandibles.txt", header = TRUE) # read data file
pairs(mandibles[, 3:11], pch=16) # print scatterplot matrix for X1 through X9
pairs(mandibles[, 3:11], col=mandibles$Group, pch=16) # coloring by Species
pairs(mandibles[, 3:11], col=1+mandibles$Sex, pch=16) # coloring by Sex
```

## (b)

I see bigger grouping differences by Species.

## (c)

```{r}
plot(1:5,1:5,col=1:5,pch=16)
```

We can see from R palette that:
* Black dots represent 1 ("Modern")
* Red dots represent group 2 ("Golden")
* Green dots represent group 3 ("Cuon")
* Dark blue dots represent 4 ("Wolf")
* Cyan dots represent 5 ("Prehistoric")

Knowing this and looking back at the pairs plot by Species we displayed in (a), we can clearly see that group 4 / "Wolf" (dark blue) has the biggest jaws and group 2 / "Golden" (red) has the smallest jaws.

## (d)

```{r}
# Filter the dataset to include only rows where Species is "Wolf"
wolf_data <- mandibles[mandibles$Species == "Wolf", ]
# Find the row with the lowest value in the X6 variable
min_row <- wolf_data[which.min(wolf_data$X6), ]
# Print the observation with the lowest X6 value for Wolf
print(min_row)
```

One dark blue point ("Wolf") looks like it is an outlier in the plot $X5$ ($length of 1st molar$) against $X6$ ($breath of 1st molar$), as it appears clearly outside the diagonal formed by other points. Subject number is 61.